Unlocking Sensitivity for Visibility-based Estimators of the 21\,cm Reionization Power Spectrum

### 2017-07-22 Rotation paper/draft_working.tex
incorporated Aaron's last round comments

Analysis pipelines for the 21\,cm power spectrum typically fall into two categories. In the first, images are formed through rotation synthesis, and after a foreground mitigation step, are Fourier transformed to construct a power spectrum \citep{MWAresult0, MWAresult1, LOFARresult}. An alternative technique works directly with visibilities from baselines, delay-transforming and cross-multiplying them to form the power spectrum \citep{delay-transform}. This technique avoids many systematics associated with combining data from different baselines and tracks the native sampling of the interferometer. An example of the visibility-based pipeline is presented in \cite{Ali2015}, which provides power spectrum
measurements with the 64-element version of PAPER (henceforth as PAPER-64). However, one disadvantage of existing visibility-based pipelines is their lack of treatment of partial redundancy. Baselines that are slightly different in length and orientation
``rotate into'' each other at a time delay. Baselines of different lengths and orientations therefore contain partially coherent information. This is the basis of earth-rotation synthesis. While imaging based power-spectrum pipelines naturally include all redundancy information, visibility-based pipelines so far only cross multiply fully redundant baselines, i.e. baselines of the same length and orientation.

@@

We begin with the visibility as commonly defined in the literature (e.g.
\citealt{TMS, first-paper}): 
\begin{equation}
\begin{aligned}V_{\nu}(\boldsymbol{b}) & =\int d\Omega A(\hat{\boldsymbol{s}},\nu)\phi(\nu)I_{\nu}(\hat{\boldsymbol{s}})\exp\left(-2\pi i\frac{\nu}{c}\boldsymbol{b}\cdot\hat{\boldsymbol{s}}\right)\\
 & \approx\frac{2k_{B}}{\lambda^{2}}\int d\Omega A(\hat{\boldsymbol{s}},\nu)\phi(\nu)T(\hat{\boldsymbol{s}})\exp\left(-2\pi i\frac{\nu}{c}\boldsymbol{b}\cdot\hat{\boldsymbol{s}}\right),
\end{aligned}
\label{eq:Vis1}
\end{equation}
where $k_B$ is Boltzmann's constant, $\lambda$ is the mean wavelength, and $\hat{\boldsymbol{s}}$ and $d \Omega$ denote a direction in the
sky and its corresponding solid angle. Inside the integral we have $\phi(\nu)$ as the frequency bandpass profile, $A(\hat{\boldsymbol{s}},\nu)$ as the (frequency-dependent) primary beam, and $I$ as the specific intensity, which has been
related to $T$, the brightness temperature in the Rayleigh-Jeans
limit. The beam power pattern $A$ is dimensionless,
normalized to 1 at its peak (zenith), and we assume it to be the same
for all baselines. Power-spectrum measurements are typically taken from $\sim10$\,MHz centered around the corresponding redshift of interest (e.g. 150\,MHz for redshift $z\approx9.5$), thus imposing a sharp drop-off on $\phi(\nu)$.

@@

Ultimately we would like to relate the observed visibility to the power spectrum, $P$, defined such that
\[
\langle \widetilde{T}^{*}(\boldsymbol{k})\widetilde{T}(\boldsymbol{k'})\rangle = \left(2\pi\right)^3\delta_D^3(\bold{k}-\bold{k'})P(k),
\label{eq:Pofk}
\]
where $\bold{k}$ is the cosmological wavenumber, $\delta_D^3$ is the three-dimensional Dirac delta function, and $\widetilde{T}(\boldsymbol{k})$ is the three-dimensional Fourier transform of the brightness temperature field $T(\bold{r})$, with $\bold r$ being the cosmological position coordinate. 
We begin by relating the observational coordinates $\hat{\boldsymbol{s}}$
and $\nu$, to
$\boldsymbol{r}$ and $\boldsymbol{k}$: 
\[
\begin{aligned}r & =\frac{c}{H_{0}}\int_{0}^{z}\frac{dz'}{E(z')}\\
 & \approx\frac{c}{H_{0}}\int_{0}^{z_{0}}\frac{dz'}{E(z')}-\frac{c(1+z_0)^{2}}{\nu_{21}H_{0}E(z_0)}\left(\nu-\nu_{0}\right)\\
 & \equiv X-Y\Delta\nu,
\end{aligned} \label{eq:r}
\]
where $r$ is the magnitude of the vector $\bold{r}$, $\nu_{21}=1420$\,MHz is the 21\,cm transition rest frequency, $\nu_{0}$ is
a reference central frequency with corresponding redshift $z_{0}$, $H_0$ is the current-day Hubble constant,  
and 
\[
E(z)=\sqrt{\Omega_{m}(1+z)^{3}+\Omega_{\Lambda}}.
\]
Inverting for $\nu$:
\begin{equation}
\nu=\frac{X-r}{Y}+\nu_{0}.\label{eq:nur}
\end{equation}

@@

\[
\begin{aligned}V_{\psi}(\boldsymbol{b'},\tau) & = \int\frac{d^{3}r}{X^{2}Y}B(\boldsymbol{\Gamma}\boldsymbol{r})T(\boldsymbol{r})\\
& \qquad \qquad  \exp\left[-2\pi i\left(\frac{\boldsymbol{b}}{c}\cdot\boldsymbol{\Gamma}\hat{\boldsymbol{r}}-\tau\right)\nu_{r}-i\psi_\nu\right],\end{aligned}
\]
where we introduced a frequency-dependent phase $\psi_{\nu}$, which to linear-order in $\nu$ corresponds to phasing the two visibilities to the same phase center. \footnote{For many cases, the linear-order interpretation is sufficient. We keep the general term to reserve the option to determine its exact form numerically. }

@@

Roughly speaking, Eqs. \eqref{eq:opp} through \eqref{eq:Thetanu} tell us that the product of visibilities at a time delay is proportional to the power spectrum times the Fourier
transform of the cross multiplied beam patterns. As a check, when applied to equivalent baselines,
$\boldsymbol{b}=\boldsymbol{b'}$, the peak correlation occurs at $\Delta t=0$ and $\hat{\boldsymbol{s}}=\boldsymbol{\Gamma}\hat{\boldsymbol{s}}$, in which case Eq. \eqref{eq:final} reduces to Eq. (B9) of \cite{paper32}. 
With Eq. \eqref{eq:opp} and Eq. \eqref{eq:Theta} we can, for any given pair of baseline classes and time delay, estimate the degree of redundancy, here represented by $\Theta$. This allows us to achieve our goals stated in the introduction: to identify 
candidate baseline pairs with significant redundancy, to find the time offset that maximizes redundancy, and to quantify the degree of such redundancy. We can do all the above simply by computing the weight $\Theta$ from
Eq. \eqref{eq:Theta} for various time offsets.

\subsection{Rephasing \label{sec:rephs}}
If $\psi_{\nu}=0$ in  Eq. \eqref{eq:Thetanu}, $\Theta_{\nu}$ at the peak of correlation is in general complex, and often far from real. Furthermore, this phase of peak correlation is inevitably frequency dependent. This frequency dependence of the phase would lead to destructive interference when we integrate over frequency, unless we correct $\Theta_{\nu}$ by an extra phase $\psi_\nu$.

To linear order in $\nu$, $\psi_\nu$ cancels the decoherence due to the two visibilities having different phase centers. By default the correlators of a drift-scan array phase the two visibilities both to zenith at the same time. When they are cross-multiplied with a time lag, the visibilities must be rephased before delay transform to account for the movement of the phase center. The effect of the phase thus roughly corresponds in a shift of the delay mode measured, and we should expect $\psi_{\nu}$ to be in first order a linear function of $\nu$:
\[
\psi_{\nu}\sim 2\pi\Delta\tau\nu,
\]
where $\Delta\tau$ is a shift in delay corresponding to the phase center movement.

In Fig. \ref{fig:phi_nu} we show the phases of $\Theta_\nu$ at the optimal time offset or, in other words, when $|\Theta_\nu|$ is maximized. The phases are shown for the baseline pair \{2,0:2,1\} of PAPER-128. The drift-scan phase dependence is compared with that of the same baseline with hypothetical tracking-elements (Eq. \eqref{eq:Thetanu} and Eq. \eqref{eq:Thetanu_tracking}). In the drift-scan case, we indeed see a linear relation corresponding to a delay of $\Delta\tau\approx15.7$\,ns. In the tracking case, since the phase center is fixed to the sky, only second order effects are observed. The origin of the second order effects can be seen as due to the $w$-term, or more precisely the fact that $uvw$ tracks do not cross when their two-dimensional projections do. We refer the reader to Appendix \ref{sec:appA} for further explanation. In both cases, the full effects are encapsulated in the phase of $\Theta_\nu$ and can be thus determined empirically without any additional computation.

@@

In this section we delve into visualizing and testing our method, and explore sensitivity contributions of various baseline classes for PAPER and HERA array configurations. 
\subsection{Visualization \label{sec:visual}}

So far for clarity and generality we have avoided the traditional formulation of radio astronomy in terms of the $uv$-plane. To gain some visual intuition of the formalism in Section \ref{sec:method}, we show explicit beam fringe patterns of the two baselines and their interactions. For this section, we relax into the flat sky approximation so that $(l,m,\sqrt{1-l^2-m^2})\equiv\hat{s}$ and examine a single frequency of $150$\,MHz.

At the time offset of maximum redundancy, we expect two baselines to have to same fringe pattern in both frequency and phase. Due to the time delay, however, the beam centers would be slightly shifted with respect to each other. In the top panels of Fig. \ref{fig:beamfringe}, we show the real parts of peak-normalized beam-fringe patterns. The top left and middle panels show the beam-fringe
patterns for the baseline classes \{2,0\} and \{2,1\}, offset by 0.78 hours. More specifically, displayed are the real parts of $g_b\equiv B(\hat{\boldsymbol{s}},\nu)\exp\left[-i2\pi\frac{\nu}{c}\hat{\boldsymbol{s}}\cdot\boldsymbol{b}\right]$ and $g_{b'}\equiv B(\boldsymbol{\Gamma}\hat{\boldsymbol{s}},\nu)\exp\left[-i2\pi\frac{\nu}{c}\boldsymbol{\Gamma}\hat{\boldsymbol{s}}\cdot\boldsymbol{b'}\right]$, respectively. The top right panel show their conjugate product $g_{b}^{*}g_{b'}e^{-i\psi_\nu} = B^{*}(\hat{\boldsymbol{s}},\nu)B(\boldsymbol{\Gamma}\hat{\boldsymbol{s}},\nu)\exp\left[i2\pi\frac{\nu}{c}\left(\hat{\boldsymbol{s}}\cdot\boldsymbol{b}-\boldsymbol{\Gamma}\hat{\boldsymbol{s}}\cdot\boldsymbol{b'}\right)-i\psi_{\nu}\right]$, where we have chosen global phase $\psi_{\nu}$ such that the peak of the conjugate product is real. The product of the beam fringe patterns in top right shows that the fringes
 cancel out as we expect. Comparing with Eq. \eqref{eq:Thetanu}, we see that $\Theta_\nu=\int d\Omega g_{b}^{*}g_{b'}$ is the sum over the image shown in the top right panel. Thus we show that at the optimal time offset the fringe patterns of the two baselines cancel out, leading to a substantial weight $\Theta_\nu$. The frequency-summed weight $\Theta$ is constructed with rephasing as described in Section \ref{sec:rephs} so that the conjugate product (top right panel) for different frequencies align in phase.

Similarly, we show the correlation of the two baselines in terms of their $uv$ beam spread overlap. The bottom left and middle panels are (peak-normalized) instantaneous $uv$ coverage of two baselines ($\widetilde{g}_{b}$ and $\widetilde{g}_{b'}$) while the bottom right panel is their product. We see from the bottom right panel that the cross-multiplied baselines recover power concentrated at $u\approx15$. The product of the $uv$ beams spread shown in the bottom right panel should not be confused with the Fourier transform of the beam-fringe conjugate product shown in the top right panel.

@@

\caption{PAPER-128 beam-fringe patterns and their point-spread functions on the $uv$-plane for a frequency of $150$\,MHz. Selected baselines have 30 meter East-West separation, and 0 and 4 meter North-South separation, respectively. The top panels show the real parts of the beam-fringe patterns of the baselines (left and middle), and their conjugate product (right). The beam-fringe values are normalized such that the peak of the original beam is unity. 
The bottom left and middle panels show the peak-normalized $uv$ beam power of the two beam-fringe patterns. The product of the beam spreads displayed in the bottom right panel shows that power concentrated at $u\approx15$ is recovered. }
\label{fig:beamfringe}
\end{figure*}

@@

Figure \ref{fig:multiblwindows} shows four example window functions. The two window functions in the top plot are for equivalent baseline power-spectrum estimators, i.e., where one squares the delay spectrum from a single baseline and normalizes the result. The window function localized at low $\ell$ is for the baseline \{2,0\} of around $30\,\textrm{m}$ in length, while the window function localized at high $\ell$ is for \{12,0\}, roughly $180\,\textrm{m}$ in length. The window functions for near-equivalent baseline estimates are shown in the bottom plot, for \{2,0:2,1\} and \{12,0:12,1\} at low- and high-$\bar{b}$, respectively. In all cases, the window functions that we show are for $\tau = 703\,\textrm{ns}$, and are generated using primary beam models for PAPER. For numerical convenience we assumed that the bandpass $\phi_\nu$ (see Eq. \eqref{eq:B}) takes the form of a half-period sine curve that peaks at $150\,\textrm{\,MHz}$ and goes to zero $5\,\textrm{\,MHz}$ on either side of the peak.

@@

To determine that actual relative contribution to sensitivity of these
baseline pairs, we also have to account for the multiplicities of
these baseline classes, i.e. the number of antenna pairs with the
same length and orientation. We would like to estimate an effective weight $\widetilde{\Theta}_{\boldsymbol{bb'}}$ that accounts for both the peak height $\Theta_{\boldsymbol{bb'}}$ and the multiplicities of the baseline classes. From Fig. \ref{fig:sensplot}
we see for example \{1,0\} has higher multiplicity than \{2,0\},
or \{1,1\}. Assuming that each equivalent baseline delivers
the same $\Theta_{\boldsymbol{bb}}$, the relative contribution to sensitivity can be estimated as follows.

First we can average the visibilities of the equivalent baselines. Since the core of PAPER-128 is a 16 by 7 antenna configuration, there
are $M\equiv(16-|m|)\times(7-|n|)$ copies of the baseline class $\{m,n\}$. This means
that if we add visibility measurements of all the equivalent baselines,
we get a factor of $\sqrt{M}$ reduction in noise
level $\sigma_N$ of the visibility. The sensitivity contribution of $\{m,n\}$, cross multiplied with $\{m',n'\}$  thus roughly scales as $\sqrt{\left((16-|m|)(7-|n|)(16-|m'|)(7-|n'|)\right)}=\sqrt{MM'}$.
For cross-multiplications of nearly equivalent baselines of
types $\{m,n\}$ and $\{m',n'\}$, we get an effective weight:

@@

Shown in the bottom panel of Fig. \ref{fig:sensplot} are the peak heights weighted
by the multiplicity factor. Points that have zero time delay are the equivalent baseline pairs and their weighted correlation values simply reflect the multiplicity factor. For clarity of presentation we have ``folded over'' the negative time delays and combined baseline pairs that are mirror images.

@@

\caption{Pairplots of the top contributing baseline pairs in three arrays. Plotted properties are optimal time delay $\Delta t_{\boldsymbol{bb'}}$,  effective weight $\widetilde{\Theta}_{\boldsymbol{bb'}}$, and baseline length $\bar{b}$. Only those points with $\widetilde{\Theta}_{\boldsymbol{bb'}}>0.01$ are shown (the weight for the top class of equivalent pair is normalized to 1). Scatter points of the three arrays overlap, in the order indicated by the legend. }
\label{fig:pairplot}
\end{figure}

@@

\section{Conclusion}
The 21\,cm brightness temperature is a direct tracer of reionization. Recently a surge of efforts in radio astronomy aims to measure the 21\,cm brightness temperature power spectrum of reionization. Measurement of the faint and diffuse signal requires high sensitivity, thus arrays are designed with multiple copies of the same baselines to enable repeated measurements of the same Fourier modes.

@@

The variance of $P$ constructed with visibilities $V_1$ and $V_2$ from two baseline classes can be estimated\footnote{We assume all noise terms to be independent for simplicity, in practice the correlation of different measurements ifrom equivalent baselines are alleviated by grouping the baselines in the class and the days of observation, as in \cite{Ali2015}}:
\begin{equation}
\begin{aligned}
\sigma_P^2 &= \langle P^2\rangle -\langle P \rangle^2,\\
&\propto \langle \frac{(V_{1S}+V_{1N})^2 (V_{2S}+V_{2N})^2}{\Theta^2} \rangle - \langle \frac{(V_{1S}+V_{1N}) (V_{2S}+V_{2N})}{\Theta} \rangle ^2,\\
&= \frac{1}{\Theta_{12}^2} \left( V_{1S}^2\sigma_{V2}^2+V_{2S}^2\sigma_{V1}^2+\langle V_{1N}^2 V_{2N}^2\rangle\right), \\
&= \frac{1}{\Theta_{12}^2} \left[ V_{S}^2(\sigma_{V2}^2+\sigma_{V1}^2) + \sigma_{V1}^2 \sigma_{V2}^2\right], 
\end{aligned}
\end{equation}
where in the second last line we have substituted in the visibility noise variance. In the final line we used Wick's theorem and the fact that the signal from the two visibilities are equal.

@@

\begin{equation}
\begin{aligned}
\sigma_P^2 & \propto  \frac{\sigma_0^4}{\Theta_{12}^2} \left[ \rho_0 \left(\frac{1}{M_1}+\frac{1}{M_2} \right) + \frac{1}{M_1 M_2}\right], \\
&\propto \frac{1}{\widetilde{\Theta}_{12}^2}.
\end{aligned}
\end{equation}
Thus we have defined a slightly modified version of the effective weight in Eq.  \eqref{eq:tildereal}:
\begin{equation}
\widetilde{\Theta}_{12}=\frac{\Theta_{12}\sqrt{M_1M_{2}}}{\sqrt{1 + \rho_0 \left(M_1+M_{2} \right)}}.
\end{equation}

>>>>

Analysis pipelines for the 21\,cm power spectrum typically fall into two categories. In the first, images are formed through rotation synthesis, and after a foreground mitigation step, are Fourier transformed to construct a power spectrum \citep{MWAresult0, MWAresult1, LOFARresult}. An alternative technique works directly with visibilities from baselines, delay-transforming and cross-multiplying them to form the power spectrum \citep{delay-transform}. This technique avoids many systematics associated with combining data from different baselines and tracks the native sampling of the interferometer. An example of the visibility-based pipeline is presented in \cite{Ali2015}, which provides power spectrum
measurements with the 64-element version of PAPER (henceforth as PAPER-64). However, one disadvantage of existing visibility-based pipelines is their lack of treatment of partial redundancy. Baselines that are slightly different in length and orientation
``rotate into'' each other at a time offset. Baselines of different lengths and orientations therefore contain partially coherent information. This is the basis of earth-rotation synthesis. While imaging based power-spectrum pipelines naturally include all redundancy information, visibility-based pipelines so far only cross multiply fully redundant baselines, i.e. baselines of the same length and orientation.

@@

We begin with the visibility as commonly defined in the literature (e.g.
\citealt{TMS, first-paper}): 
\begin{equation}
\begin{aligned}V_{\nu}(\boldsymbol{b}) & =\int d\Omega A(\hat{\boldsymbol{s}},\nu)\phi(\nu)I_{\nu}(\hat{\boldsymbol{s}})\exp\left(-2\pi i\frac{\nu}{c}\boldsymbol{b}\cdot\hat{\boldsymbol{s}}\right)\\
 & \approx\frac{2k_{B}}{\lambda^{2}}\int d\Omega A(\hat{\boldsymbol{s}},\nu)\phi(\nu)T(\hat{\boldsymbol{s}})\exp\left(-2\pi i\frac{\nu}{c}\boldsymbol{b}\cdot\hat{\boldsymbol{s}}\right),
\end{aligned}
\label{eq:Vis1}
\end{equation}
where $k_B$ is Boltzmann's constant, $\lambda$ is the mean wavelength, and $\hat{\boldsymbol{s}}$ and $d \Omega$ denote a direction in the
sky and its corresponding solid angle. Inside the integral we have $\phi(\nu)$ as the frequency bandpass profile, $A(\hat{\boldsymbol{s}},\nu)$ as the (frequency-dependent) primary beam, and $I$ as the specific intensity, which has been
related to $T$, the brightness temperature in the Rayleigh-Jeans
limit. The beam power pattern $A$ is dimensionless,
normalized to 1 at its peak (zenith), and we assume it to be the same
for all baselines. Power-spectrum measurements are typically taken from $\sim0.01$\,GHz centered around the corresponding redshift of interest (e.g. 0.15\,GHz for redshift $z\approx9.5$), thus imposing a sharp drop-off on $\phi(\nu)$.

@@

Ultimately we would like to relate the observed visibility to the power spectrum, $P$, defined such that
\[
\langle \widetilde{T}^{*}(\boldsymbol{k})\widetilde{T}(\boldsymbol{k'})\rangle = \left(2\pi\right)^3\delta_D^3(\bold{k}-\bold{k'})P(k),
\label{eq:Pofk}
\]
where $\bold{k}$ is the cosmological wavenumber, $\delta_D^3$ is the three-dimensional Dirac delta function, and $\widetilde{T}(\boldsymbol{k})$ is the three-dimensional Fourier transform of the brightness temperature field $T(\bold{r})$, with $\bold r$ being the cosmological position coordinate. 
We begin by relating the observational coordinates $\hat{\boldsymbol{s}}$
and $\nu$, to
$\boldsymbol{r}$ and $\boldsymbol{k}$: 
\[
\begin{aligned}r & =\frac{c}{H_{0}}\int_{0}^{z}\frac{dz'}{E(z')}\\
 & \approx\frac{c}{H_{0}}\int_{0}^{z_{0}}\frac{dz'}{E(z')}-\frac{c(1+z_0)^{2}}{\nu_{21}H_{0}E(z_0)}\left(\nu-\nu_{0}\right)\\
 & \equiv X-Y\Delta\nu,
\end{aligned} \label{eq:r}
\]
where $r$ is the magnitude of the vector $\bold{r}$, $\nu_{21}=1.420$\,GHz is the 21\,cm transition rest frequency, $\nu_{0}$ is
a reference central frequency with corresponding redshift $z_{0}$, $H_0$ is the current-day Hubble constant,  
and 
\[
E(z)=\sqrt{\Omega_{m}(1+z)^{3}+\Omega_{\Lambda}}.
\]
Inverting for $\nu$:
\begin{equation}
\nu=\frac{X-r}{Y}+\nu_{0}.\label{eq:nur}
\end{equation}

@@

\[
\begin{aligned}V_{\psi}(\boldsymbol{b'},\tau) & = \int\frac{d^{3}r}{X^{2}Y}B(\boldsymbol{\Gamma}\boldsymbol{r})T(\boldsymbol{r})\\
& \qquad \qquad  \exp\left[-2\pi i\left(\frac{\boldsymbol{b}}{c}\cdot\boldsymbol{\Gamma}\hat{\boldsymbol{r}}-\tau\right)\nu_{r}-i\psi_\nu\right],\end{aligned}
\]
where we introduced a frequency-dependent phase $\psi_{\nu}$, which to linear-order in $\nu$ corresponds to phasing the two visibilities to the same phase center. \footnote{For many cases, the linear-order interpretation is sufficient. We keep the general term to reserve the option to determine its exact form numerically as this information comes at no additional computational cost. }

@@

Roughly speaking, Eqs. \eqref{eq:opp} through \eqref{eq:Thetanu} tell us that the product of visibilities at a time offset is proportional to the power spectrum times the Fourier
transform of the cross multiplied beam patterns. As a check, when applied to equivalent baselines,
$\boldsymbol{b}=\boldsymbol{b'}$, the peak correlation occurs at $\Delta t=0$ and $\hat{\boldsymbol{s}}=\boldsymbol{\Gamma}\hat{\boldsymbol{s}}$, in which case Eq. \eqref{eq:final} reduces to Eq. (B9) of \cite{paper32}. 
With Eq. \eqref{eq:opp} and Eq. \eqref{eq:Theta} we can, for any given pair of baseline classes and time offsets, estimate the degree of redundancy, here represented by $\Theta$. This allows us to achieve our goals stated in the introduction: to identify 
candidate baseline pairs with significant redundancy, to find the time offset that maximizes redundancy, and to quantify the degree of such redundancy. We can do all the above simply by computing the weight $\Theta$ from
Eq. \eqref{eq:Theta} for various time offsets.

\subsection{Rephasing \label{sec:rephs}}
If $\psi_{\nu}=0$ in  Eq. \eqref{eq:Thetanu}, $\Theta_{\nu}$ at the peak of correlation ( as a function of time) is generally complex, and often far from real. Furthermore, this phase of peak correlation is inevitably frequency dependent. This frequency dependence would lead to destructive interference when we integrate over frequency, unless we correct $\Theta_{\nu}$ by a phase $\psi_\nu$.

To linear order in $\nu$, $\psi_\nu$ cancels the decoherence due to the two visibilities having different phase centers. By default the correlators of a drift-scan array phase the two visibilities both to zenith at the same time. When they are cross-multiplied with a time lag, the visibilities must be rephased before the delay transform to account for the movement of the phase center. The effect of the phase thus roughly corresponds in a shift of the delay mode measured, and we should expect $\psi_{\nu}$ to be approximately a linear function of $\nu$:
\[
\psi_{\nu}\sim 2\pi\Delta\tau\nu,
\]
where $\Delta\tau$ is a shift in delay corresponding to the phase center movement.

In Fig. \ref{fig:phi_nu} we show the phases of $\Theta_\nu$ at the optimal time offset, i.e. when $|\Theta_\nu|$ is maximized. The phases are shown for the baseline pair \{2,0:2,1\} of PAPER-128. The drift-scan phase dependence is compared with that of the same baseline with hypothetical tracking-elements (Eq. \eqref{eq:Thetanu} and Eq. \eqref{eq:Thetanu_tracking}). In the drift-scan case, we indeed see a linear relation corresponding to a delay of $\Delta\tau\approx15.7$\,ns. This delay corresponds to the light travel time from one antenna to another of the rotated baseline. In the tracking case, since the phase center is fixed to the sky, only second order effects are observed. The origin of the second order effects can be seen as due to the $w$-term, or more precisely the fact that $uvw$ tracks do not cross when their two-dimensional projections do. We refer the reader to Appendix \ref{sec:appA} for further explanation. In both cases, the full effects are encapsulated in the phase of $\Theta_\nu$ and can be thus determined empirically without any additional computation.

@@

In this section we delve into visualizations and analyses of our method, and explore sensitivity contributions of various baseline classes for PAPER and HERA array configurations. 
\subsection{Visualization \label{sec:visual}}

@@

\caption{PAPER-128 beam-fringe patterns and their point-spread functions on the $uv$-plane for a frequency of $0.15$\,GHz. Selected baselines have 30-meter East-West separations, and 0- and 4-meter North-South separations, respectively. The top panels show the real parts of the beam-fringe patterns of the baselines (left and middle), and their conjugate product (right). The beam-fringe values are normalized such that the peak of the original beam is unity. 
The bottom-left and middle panels show the peak-normalized $uv$ point spread of the two beam-fringe patterns. The product of the beam spreads displayed in the bottom-right panel shows that power concentrated at $u\approx15$ is recovered. }
\label{fig:beamfringe}
\end{figure*}

So far for clarity and generality we have avoided the traditional formulation of radio astronomy in terms of the $uv$-plane. To gain some visual intuition of the formalism in Section \ref{sec:method}, we show explicit beam fringe patterns of the two baselines and their interactions. For this section, we relax into the flat sky approximation so that $(l,m,\sqrt{1-l^2-m^2})\equiv\hat{s}$ and examine a single frequency of $0.15$\,GHz.

At the time offset of maximum redundancy, we expect two baselines to have to same fringe pattern in both frequency and phase. Due to the time offset, however, the beam centers would be slightly shifted with respect to each other. In the top panels of Fig. \ref{fig:beamfringe}, we show the real parts of peak-normalized beam-fringe patterns. The top left and middle panels show the beam-fringe
patterns for the baseline classes \{2,0\} and \{2,1\}, offset by 0.78 hours. More specifically, displayed are the real parts of $g_b\equiv B(\hat{\boldsymbol{s}},\nu)\exp\left[-i2\pi\frac{\nu}{c}\hat{\boldsymbol{s}}\cdot\boldsymbol{b}\right]$ and $g_{b'}\equiv B(\boldsymbol{\Gamma}\hat{\boldsymbol{s}},\nu)\exp\left[-i2\pi\frac{\nu}{c}\boldsymbol{\Gamma}\hat{\boldsymbol{s}}\cdot\boldsymbol{b'}\right]$, respectively. The top right panel show their conjugate product $g_{b}^{*}g_{b'}e^{-i\psi_\nu} = B^{*}(\hat{\boldsymbol{s}},\nu)B(\boldsymbol{\Gamma}\hat{\boldsymbol{s}},\nu)\exp\left[i2\pi\frac{\nu}{c}\left(\hat{\boldsymbol{s}}\cdot\boldsymbol{b}-\boldsymbol{\Gamma}\hat{\boldsymbol{s}}\cdot\boldsymbol{b'}\right)-i\psi_{\nu}\right]$, where we have chosen global phase $\psi_{\nu}$ such that the peak of the conjugate product is real. The product of the beam fringe patterns in the top right shows that the fringes
 cancel out as we expect. Comparing with Eq. \eqref{eq:Thetanu}, we see that $\Theta_\nu=\int d\Omega g_{b}^{*}g_{b'}e^{-i\psi_\nu}$ is the sum over the image shown in the top right panel. At the optimal time offset, the fringe patterns of the two baselines cancel out, leading to a substantial weight $\Theta_\nu$. The frequency-summed weight $\Theta$ is constructed with rephasing as described in Section \ref{sec:rephs} so that the conjugate product (top-right panel) for different frequencies align in phase.

Similarly, we show the correlation of the two baselines in terms of the overlap of their point-spread functions in the $uv$-plane. The bottom-left and middle panels are the peak-normalized instantaneous $uv$ coverage of two baselines ($\widetilde{g}_{b}$ and $\widetilde{g}_{b'}$) while the bottom right panel is their product. We see from the bottom right panel that the cross-multiplied baselines' recovered power is concentrated at $u\approx15$. The product of the $uv$ point-spread functions shown in the bottom-right panel should not be confused with the Fourier transform of the beam-fringe conjugate product shown in the top-right panel.

@@

Figure \ref{fig:multiblwindows} shows four example window functions. The two window functions in the top plot are for equivalent baseline power-spectrum estimators, i.e., where one squares the delay spectrum from a single baseline and normalizes the result. The window function localized at low $\ell$ is for the baseline \{2,0\} of around $30\,\textrm{m}$ in length, while the window function localized at high $\ell$ is for \{12,0\}, roughly $180\,\textrm{m}$ in length. The window functions for near-equivalent baseline estimates are shown in the bottom plot, for \{2,0:2,1\} and \{12,0:12,1\} at low- and high-$\bar{b}$, respectively. In all cases, the window functions that we show are for $\tau = 703\,\textrm{ns}$, and are generated using primary beam models for PAPER. For numerical convenience we assumed that the bandpass $\phi_\nu$ (see Eq. \eqref{eq:B}) takes the form of a half-period sine curve that peaks at $0.15\,\textrm{\,GHz}$ and goes to zero $5\,\textrm{\,MHz}$ on either side of the peak.

@@

To determine that actual relative contribution to sensitivity of these
baseline pairs, we also have to account for the multiplicities of
these baseline classes, i.e. the number of antenna pairs with the
same length and orientation. We would like to estimate an effective weight $\widetilde{\Theta}_{\boldsymbol{bb'}}$ that accounts for both the peak height $\Theta_{\boldsymbol{bb'}}$ and the multiplicities of the baseline classes. From Fig. \ref{fig:sensplot}
we see for example that \{1,0\} has higher multiplicity than \{2,0\},
or \{1,1\}. Assuming that each equivalent baseline delivers
the same $\Theta_{\boldsymbol{bb}}$, the relative contribution to sensitivity can be estimated as follows.

First we can average the visibilities of the equivalent baselines. Since the core of PAPER-128 is a 16 by 7 antenna configuration, there
are $M\equiv(16-|m|)\times(7-|n|)$ copies of the baseline class $\{m,n\}$. This means
that if we add visibility measurements of all the equivalent baselines,
we get a factor of $\sqrt{M}$ reduction in the noise
level $\sigma_N$ of the visibility. The sensitivity contribution of $\{m,n\}$, cross multiplied with $\{m',n'\}$  thus roughly scales as $\sqrt{\left((16-|m|)(7-|n|)(16-|m'|)(7-|n'|)\right)}=\sqrt{MM'}$.
For cross-multiplications of nearly equivalent baselines of
types $\{m,n\}$ and $\{m',n'\}$, we get an effective weight:

@@

Shown in the bottom panel of Fig. \ref{fig:sensplot} are the peak heights weighted
by the multiplicity factor. Points that have zero optimal time offset are the equivalent baseline pairs and their weighted correlation values simply reflect the multiplicity factor. For clarity of presentation we have ``folded over'' the negative time offsets and combined baseline pairs that are mirror images.

@@

\caption{Pairplots of the top contributing baseline pairs in three arrays. Plotted properties are optimal time offset $\Delta t_{\boldsymbol{bb'}}$ (in hours),  effective weight $\widetilde{\Theta}_{\boldsymbol{bb'}}$, and baseline length $\bar{b}$ (in meters). Only those points with $\widetilde{\Theta}_{\boldsymbol{bb'}}>0.01$ are shown (the weight for the top class of equivalent pair is normalized to 1). Scatter points of the three arrays overlap, in the order indicated by the legend. }
\label{fig:pairplot}
\end{figure}

@@

\section{Conclusion}
Upcoming arrays aiming at measuring the 21\,cm power spectrum, including HERA and SKA-low, feature increasing number of antennas on regular grids to maximize coherent information. Having efficient pipelines to extract full sensitivity of these arrays is thus a pressing necessity.

@@

The variance of $P$ constructed with visibilities $V_1$ and $V_2$ from two baseline classes can be estimated\footnote{We assume all noise terms to be independent for simplicity, in practice the correlation of different measurements ifrom equivalent baselines are alleviated by grouping the baselines in the class and the days of observation, as in \cite{Ali2015}}:
\begin{equation}
\begin{aligned}
\sigma_P^2 &= \langle P^2\rangle -\langle P \rangle^2,\\
&\propto \langle \frac{(V_{1S}+V_{1N})^2 (V_{2S}+V_{2N})^2}{\Theta^2} \rangle - \langle \frac{(V_{1S}+V_{1N}) (V_{2S}+V_{2N})}{\Theta} \rangle ^2\\
&= \frac{1}{\Theta_{12}^2} \left( V_{1S}^2\sigma_{V2}^2+V_{2S}^2\sigma_{V1}^2+\langle V_{1N}^2 V_{2N}^2\rangle\right)\\
&= \frac{1}{\Theta_{12}^2} \left[ V_{S}^2(\sigma_{V2}^2+\sigma_{V1}^2) + \sigma_{V1}^2 \sigma_{V2}^2\right], 
\end{aligned}
\end{equation}
where in the second last line we have substituted in the visibility noise variance. In the final line we used Wick's theorem and the fact that the signal from the two visibilities are equal.

@@

\begin{equation}
\begin{aligned}
\sigma_P^2 & \propto  \frac{\sigma_0^4}{\Theta_{12}^2} \left[ \rho_0 \left(\frac{1}{M_1}+\frac{1}{M_2} \right) + \frac{1}{M_1 M_2}\right]\\
&\propto \frac{1}{\widetilde{\Theta}_{12}^2}.
\end{aligned}
\end{equation}
Thus we have defined a slightly modified version of the effective weight in Eq.  \eqref{eq:tildereal}:
\begin{equation}
\widetilde{\Theta}_{12}=\frac{\Theta_{12}\sqrt{M_1M_{2}}}{\sqrt{1 + \rho_0 \left(M_1+M_{2} \right)}}.
\end{equation}
